---
title: "Module 3 - Hypothesis Testing"
output: 
  learnr::tutorial:
    css: css/style.css
runtime: shiny_prerendered
description: > 
  Hypothesis Testing Module for the 2025 Envision course
editor_options: 
  markdown: 
    wrap: 72
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(tidyverse)
library(shiny)
library(gt)
library(plotly)
library(png)
library(jpeg)
library(grid)
library(cowplot)

Whale<-read.csv("Data/Humpback.csv") %>% filter(length.m.>0) %>% filter(Year>=1949 & Year <=1965)
Whale$sex<-factor(Whale$sex, levels = c("Male", "Female", "Hermaphrodite", "Unknown"))

Whale_62<-filter(Whale,Year==1962)


Whale_49<-filter(Whale,Year==1949)

choices<-c("Boxplot"="boxplot",
"Histogram"="histogram",
"Density Plot"="density")


plot_theme<-theme_light()+
  theme(plot.title=element_text(face =2,size=18),
        axis.title=element_text(face =2,size=14),
        plot.subtitle=element_text(face =3,size=14),
        axis.text=element_text(face =1,size=12),
        strip.text=element_text(face =1,size=12))


Whale %>%
  ungroup() %>%
      summarise("Number of whales"=n(),
                "Mean length"=mean(length.m.)%>%round(1),
                "Median length"=median(length.m.)%>%round(1),
                "Minimum"=min(length.m.)%>%round(1),
                "Maximum"=max(length.m.)%>%round(1),
                "Range"=(max(length.m.)-min(length.m.))%>%round(1),
                "Standard Deviation"=sd(length.m.)%>%round(1),
                "Variance"=var(length.m.)%>%round(),
                )%>%
  gather(Statistic,Value) ->table_stat
```

## Introduction

In this third workbook, we are going to be taking a closer look at hypothesis testing.
 
Hypothesis testing is the process that allows us to answer questions about our data, our population and our models. It is a well defined process that is performed always in the same way, although the calculations and the data involved will differ depending on the test performed or the question to be answered. It is related to and can used along some of the results we learned in Module 2, especially the concept of precision measures such as the confidence interval. And we will use the method of hypothesis testing in Module 4 and 5 when we will discuss model fitting.

This is a basic topic in statistics, because it lays the foundations of the *statistical inference*, the process that allows us to make statements and answer questions about a population with incomplete information — we only have the data contained in the sample. 

The experience tells us that this is one of the hardest topics to learn. We’ll try to avoid all the mathematical apparatus around it, and concentrate in a few important concepts and in the process to test statistical hypothesis.

In this session, we will cover:

-   What is a statistical hypothesis and the concept of the null hypothesis.

-   The concept of p-value, whether to use one side or two sides of the sampling distribution to compute it, and how it is related to the confidence interval seen in Module 2.

-   What is the statistical significance, α.

-   The process of testing hypothesis. 

-   Some basic statistical tests that use the hypothesis testing method.

In these workbooks all output produced will come from the use of R, and all of the code used in the workbooks can be found on GitHub here: https://github.com/stats4sd/Stats_ReIntro_M3. But the principles are identical regardless of which data analysis tool is being used. Whichever tool is being used for data analysis - it will require that the user is familiar enough with it to be able to generate any sort of summary or plot or transformation that they might suddenly think could be useful!


## Data being used in this workbook
In this session we are going to use a public dataset produced by the Ocean Biodiversity Information system (OBIS). It contains the Historic whale catch records from Australian whaling stations (1925-1978). The full data set and other details can be found here <https://obis.org/dataset/43e4c2cc-5886-427f-a4a9-d47d2635d246>. It contains 38267 records, but we will use a subset: the data corresponding to the most common caught whale, the Humpback (Megaptera novaeangliae) in the years 1949 to 1962. 

We have also edited it to contain only the columns we are interested in. Below you can explore the the data we are going to use in this workbook.

```{r, echo=FALSE}
DT::datatable(Whale)
```

## What is an 'Hypothesis' 

**Question**: What is the proportion of males among humpback whales? 

Without knowing too much about whales, the obvious answer might be to guess that it would be 50%. Actually within our dataset there was one labelled as a hermaphrodite, not enough to make much of a difference to the % - but we will consider the variable of interest here as "the % of whales which are male", for the sake of creating an easier to analyse binary variable.

But let's have a look at the data to see the results. The figure below shows the percentage over the period 1949-1962. Within each year there were we find at least 56% of the whales were male, with the figure as high as 80% in 1951.


```{r fig1, out.width="80%", echo = FALSE}
male_year<-Whale %>%
  group_by(Year) %>%
  summarise(Prop_male=sum(sex=="Male")/n(),n=n())%>%
  filter(Year>=1945 & Year<=1962)

(ggplot(male_year, aes(x=Year,y=Prop_male))+
  geom_line()+
  geom_point(aes(size=n))+
  geom_smooth(linewidth = 0.5,colour = "red", se = FALSE) +
  geom_line(aes(y=0.5), linetype=2)+
  coord_cartesian(ylim = c(0.25,0.90))+
  labs(x="Year",y="% Male")+
    scale_x_continuous(breaks=seq(1950,1964,by=2))+
    ggtitle("Whales observed in Australia over time: % Male")+
    scale_size_continuous(range=c(2,4))+
  scale_y_continuous(labels=scales::percent,breaks=seq(0,1,by=0.1))+
  plot_theme) %>%
  ggplotly()
```

If there were no difference in the underlying proportions of male and female whales, then each year there would be a 50:50 chance of having more male than female catches, or more female than male catches - just like flipping a coin. 

However, in this data we see that every year we have observed more male than female. If it were indeed a 50:50 gender ratio then this would be like flipping a fair coin 14 times in a row and getting heads every time. The probability of this happening is 0.5 raised to the power 14, which is a very small number: 0.0000610.

If we observed this in a coin-flipping experiment, we could confidently claim that the coin was not fair.

So, we can ask ourselves whether the true proportion of males among the humpback whales can really be 50%? The process we follow to answer a question like this is called *hypothesis testing*.

We can define a hypothesis as a proposed explanation for a phenomenon. It is not the absolute truth, but rather a provisional, working assumption. Within a hypothesis test we look at the evidence against that hypothesis, by comparing how consistent our data is with that hypothesis. 


## The Null Hypothesis

The *null hypothesis* ($H_0$) is the 'default' hypothesis that we work with until we have sufficient evidence against it, and it generally is used as a hypothesis to indicated that 'nothing' is happening, i.e. there are no trends or no differences between groups or the value of interest is equal to 0. Hence the word "null". In the example of the whales, we could state the null hypothesis as:

-   The proportion of male whales is 50% ($H_0: p_M = 0.5$) or

-   The proportion of male whales is equal to the proportion of female whales ($H_0: p_M = p_F$)

The null hypothesis is what we are willing to assume is the case, until proven otherwise. It is extremely conservative, denying any progress or change! However, this does not mean that we believe the null hypothesis is literally true; we are never looking to claim that the null hypothesis has actually been proved; instead we are only ever looking for evidence *against* the null hypothesis. 

In the words of Ronald Fisher, the father of modern statistics:

*"the null hypothesis is never proved or established, but is possibly disproved, in the course of experimentation. Every experiment may be said to exist only in order to give the facts a chance of disproving the null hypothesis."*

*(R.A. Fisher (1935), The Design of Experiments. Oliver and Boyd, p.19).*

There is a strong analogy with criminal trials in the English – and indeed many other – legal systems: a defendant can be found guilty, but nobody is ever found innocent, simply not proven guilty. Similarly, we may find that we reject the null hypothesis, but if we don't have enough evidence to do so, it doesn't mean that we can accept it as truth. It is just a working assumption until something better comes along.  


### The alternative hypothesis

The alternative hypothesis ($H_1$) is the statement that is being tested against the null hypothesis. It claims that there is an effect in the population that is caused by a non-random cause. *When we have enough evidence to against the null hypothesis, we reject it, and we accept the alternative being true.*

Often, the alternative hypothesis is the same as the research hypothesis. In other words, it’s the claim that you expect or hope will be true.

The alternative hypothesis is the complement to the null hypothesis. Null and alternative hypotheses are exhaustive, meaning that together they cover every possible outcome. They are also mutually exclusive, meaning that only one can be true at a time.

Alternative hypotheses often include phrases such as “an effect”, “a difference”, or “a relationship”. When alternative hypotheses are written in mathematical terms, they always include an inequality (usually ≠, but sometimes > or <). As with null hypotheses, there are many acceptable ways to phrase an alternative hypothesis.

In the example of the whales, we could state the null hypothesis as: the proportion of male whales is 50%, or the proportion of male whales is equal to the proportion of female whales. The alternative hypothesis, then, will be one of the three following options:  

-    $H_1: p_M \ne 0.5$ or $H_1: p_M \ne p_F$

-    $H_1: p_M \gt 0.5$ or $H_1: p_M \gt p_F$

-    $H_1: p_M \lt 0.5$ or $H_1: p_M \lt p_F$



## The p-value

Let's go just to the final year of our data, 1962, and take a look at the results:

```{r}
Whale %>%
  filter(Year==1962) %>%
    group_by(Year,sex) %>%
      summarise(n=n()) %>%
        mutate(percentage=scales::percent(n/sum(n),0.1)) %>%
  ungroup() %>%
  gt()
```

We observed 720 whales in this year, of which 409, 56.8%, were male. 

This feels like quite a large difference - but a crucial question is whether this difference is large enough to provide evidence against the null hypothesis, or whether we might expect to sometimes see such a difference purely based on chance. 

If our null hypothesis were true then the probability of any randomly observed whale being male would be 50%, the same as obtaining heads when tossing a coin. But across 720 coin flips we wouldn't expect to always observe exactly 50% heads and 50% tails - sometimes there would be more and sometimes less.

We could run a simulation of 10,000 experiments where we did 720 coin flips to get an idea of the distribution of how much variability around 50% we would expect to see occurring purely by chance.


```{r fig2, out.width="80%", echo = FALSE}
set.seed(216486)

heads<-data.frame(Males=rbinom(10000,720,0.5))

pl1<-ggplot(heads, aes(x=Males))+
  geom_bar()+
  geom_vline(xintercept = 409, linetype=2,colour = "red")+
  labs(x="Number of Males",y="Number of Simulations")

ggplotly(pl1+
  plot_theme)

table(heads>=409)

```

Out of the 10,000 simulations, assuming equal proportions of males and females, we saw an outcome 409 or more 'males' only 4 times. 

So, is it possible that, if the true probability of male and female whales is equal (0.5), we could observed 409 males in a random sample of 720 whales? The answer is YES, it is possible, although very unlikely - in this simulation it happened just 0.04% of the time.

Instead of running a simulation of 10,000 runs, we could have systematically worked through all the possible permutations of the sample and calculated a theoretical probability under the null hypothesis. Each of these would generate an observed proportion of male whales, and plotting these would produce a smoother distribution than just using 10,000 simulations. Unfortunately, the number of permutations is so high that it cannot be obtained with current means within a person's lifetime.

Fortunately, though, we don’t have to perform these calculations, since the probability distribution for the proportions under the null hypothesis is based on the binomial distribution. From this we can determine the probability of a number of successes out of n trials for a given probability of success — 0.5 in our case.

The figure below shows the probability density function (A) and the cumulative probability distribution function (B). In both cases, the value 409 is highlighted with a vertical line, to show how likely it is to obtain 409 successes or more in 720 trials where the probability of success is 0.5. This probability is 0.00011, or 0.011%.
     

```{r fig3, out.width="80%", echo = FALSE}

tots<- seq(1:720)
binomial.df<-dbinom(tots,720,0.5)
cumbinomial.df<-pbinom(tots,720,0.5)

distr.ft<-data.frame(tots, binomial.df, cumbinomial.df)

pl2 <- ggplot(distr.ft, aes(x=tots, y=binomial.df))+
  geom_line()+
  coord_cartesian(xlim = c(300,420))+
  labs(x="Num. of Males",y="Proportion")+
  geom_vline(xintercept = 409, linetype=2,colour = "red")

pl2


pl3 <- ggplot(distr.ft, aes(x=tots, y=cumbinomial.df))+
  geom_line()+
  coord_cartesian(xlim = c(300,420))+
  labs(x="Number of Males",y="Cumulative Proportion")+
  geom_vline(xintercept = 409, linetype=2,colour = "red")

pl3

plot_grid(pl2, pl3, labels = c('A', 'B'), label_size = 12)
```

This probability is the so-called *p-value*, one of the most prominent concepts in statistics as practised today. A more formal definition of *p-value* is *a p-value is the probability of obtaining a result at least as extreme as the one we observed assuming the null hypothesis is really true*. As this p-value starts to decrease the strength of evidence against the null hypothesis starts to increase - it is starting to seem like our observations are not compatible with our null hypothesis.


## One-sided or two-sided test

The p-value is the probability of obtaining a result at least as extreme as the observed result if the null hypothesis were really true. What do we mean by *extreme*? 

The p-value we obtained, 0.011% is *one-tailed*, corresponding to what we know as *one-sided test*, the test that uses as alternative hypothesis $H_1: p_M \gt 0.5$. Because we are only considering only how likely is that we have observed such an extreme value (56.8%) where males were more common than females. 

However, if we had observed the reverse - and our data contained whales of which 56.8% were female, then it would also have made us suspect that the null hypothesis does not hold. Therefore, we should also calculate the probability of observing 56.8% or more of either male or female whales. This is what is known as the *two-tailed* p-value that corresponds to the *two-sided test*. This test uses as alternative hypothesis $H_1: p_M \ne 0.5$.
This two-tail probability is double the one-tailed probability, 0.021% or 0.0022, since the binomial distribution is perfectly symmetrical as we saw in the previous section.


```{r, out.width="80%", echo = FALSE}
pl2+
  geom_vline(xintercept = 720-409, linetype=2,colour = "red")+
  plot_theme

```



## Statistical significance

In our example of the proportion of male and female whales, we had the following:

-   *Null hypothesis*: the proportion of male whales is 50%.

-   *Alternative hypothesis*: the proportion of male whales is **not equal** to 50%.

-   We obtained a sample of *n=720* whales.

-   We observed the sex and *measured the proportion of male whales in the sample: 56.8%*.

-   We obtained the *p-value = 0.00022*: the probability of obtaining a value as extreme as this if the null hypothesis is true. This is the *two-sided test*.

While there is a very small chance that we could have obtained this percentage if the null hypothesis was really true, it seems very unlikely. So the question is know whether we think the *p-value* is small enough for us to be able to consider that the difference is *statistically significant*, and, therefore, the alternative hypothesis true.

In other words, whether the difference in our estimate the difference between the population value (the proportion of males) and the null hypothesis is *not equal to zero*. 

It seems extremely plausible that the difference between the proportion of males observed in the sample and the null hypothesis value cannot be explained by random variation alone. It therefore must have occurred because the population value *is not equal to the null hypothesis*, and the sample data provides the evidence for this. 

Since we have incomplete information about the population — we only have the sample — we can always be wrong when making statements like like the following: *"the proportion of male whales is significantly different from 0.5"*. However, we control the risk of being wrong when rejecting the null hypothesis through the p-value. 


Using the analogy of criminal trials in the English legal system, the sample data provide the evidence against the innocence of the defendant, and p-value is the probability of being wrong if the defendant is declared guilty. Even if we had observed 720 male whales there will still be a small (extremely small!) probability of that happening where the proportion of males was genuinely 50%.

So we need to define a *significance level*, denoted by α, which is the critical threshold value for the p-value below which we deem that we have sufficient evidence in our case against the null hypothesis. Therefore: 

-   if the p-value ≤ α, the result is statistically significant, and we reject the null hypothesis
 
-   if the p-value > α, the result is not statistically significant, and we cannot reject the null hypothesis

The most commonly used significance level (α) is 0.05 or 5%, but in some fields a level of 0.1 (10%) or 0.01 (1%) is more conventional, and there is a strong movement in favour of setting the threshold at 0.005 (https://www.nature.com/articles/s41562-017-0189-z). Therefore, some hypothetical results may be significant at one threshold but not at an other.

However, we must acknowledge that this threshold is arbitrary. For example, if we use α = 0.05, a p-value of 0.0499 is essentially not different from a p-value of 0.0501, but we would make a completely different decision if we strictly followed the significance level rule. P-values well above or well below α lead us to a straightforward decision, whereas p-values close to α usually lead to an inconclusive decision.



## Hypothesis testing

There are a large number of different 'hypothesis tests' (see the section on *Statistical tests*), each of which allows us to address a slightly different set of hypothesis. But the general process in all of them follows the same steps: 

1-  Set up a question in terms of a null hypothesis and alternative hypothesis that we want to check. We usually use the notation $H_0$ for the null hypothesis and $H_1$ for the alternative hypothesis. 

2-  Choose a test statistic that estimates something that, if it turns out to be extreme enough, would lead us to doubt the null hypothesis. You will not actually have to find this statistic, as each (standard) statistical test has already defined this test statistic.  

3-  Generate the sampling distribution of this test statistic, if the null hypothesis was true. 

4-  Obtain the p-value comparing your tests statistic with the sampling distribution. The p-value is the probability of observing such an extreme statistic if the null hypothesis is true. We have to decide if we are going to use the one-sided or two-sided p-value. In most of the tests, we will use the two-sided p-value. 

5-  Declare the result is significant if the p-value is below some critical threshold α.

There is a good chance if you have been taught statistics at school or undergraduate level you may have had to calculate many of these steps by hand, and been asked to consult long tables of numbers containing test statistics and critical thresholds. 

Thankfully this process has been obsolete for at least 50 years at this point (even if it is still regularly taught to students!), as we now have computers to do this for us.

It is important to emphasise that the exact p-value is conditional not only on the truth of the null hypothesis, but also on all other assumptions underlying the statistical model, such as lack of systematic bias, independence of the observations, the distribution of the variables. We'll discuss some of these assumptions later.


### Confidence Interval for hypothesis testing

Although the confidence interval is a tool for estimation, we can use it also for hypothesis testing. Instead of generating the sampling distribution under the null hypothesis for the tests statistic and comparing the tests statistic with this sampling distribution to obtain the p-value (steps 3 & 4 in the hypothesis testing process above), we generate the confidence interval and observe whether the null hypothesis value lies within its limits or outside them. Then, if the null hypothesis value lies within the confidence interval limits, we can't reject it, the results are not significant, whereas if the value lies outside the confidence interval limits, the result is significant and there is sufficient evidence to reject the null hypothesis.

Using a confidence level of 95% and a significance level of 5% will lead to the same results, because they are complementary. And the same happens for a confidence level of 99% and a significance level of 1%. If a 99% confidence interval does not include the null hypothesis, then we can reject it with a p-value < 0.01.

ADD SOME SORT OF PLOT HERE TO ILLUSTRATE?


## Statistical tests

In this section, we are going to discuss some of the most commonly taught tests that could be used to answer questions about our data. We are not going to go into the mathematical details, we will focus on the type of test and things to consider when running the test. There are plenty of resources on this tests that you can easily find. (ADD IN THESE RESOURCES)

Before any formal hypothesis test, always explore your data look at summary statistics and plots showing the relationship you are interested in.


### Comparing the mean of one numeric variable to a specific value: t-test


The *one sample t-test* tests the null hypothesis that the population mean is equal to a specified value: $$H_0: \mu = \mu_0$$, where $\mu_0$ is a numeric value. 
the test statistic is $$ t=\frac{\bar{x}- \mu_0}{\frac{s}{\sqrt{n}}}$$, where
 
-   $\bar{x}$ is the sample mean
 
-   $\mu_0$ is the null hypothesis value to be tested
 
-    *s* is the sample standard deviation
 
-    *n* is the sample size.

This *t* statistic is compared to its sampling distribution were the null hypothesis true (a *Student's t* with *n-1* degrees of freedom), and a p-value is obtained. Then a decision is made based on the p-value and the level of significance we wish to set. 

Lets look at our whales from a different year to before, this time 1949. The summary statistics are:
 
```{r}
Whale_49 %>%
    group_by(Year) %>%
      summarise(n=n(),"Mean Length"=mean(length.m.),"SD Length"=sd(length.m.)) %>%
  ungroup() %>%
  gt()%>%
    fmt_number(columns =  3:4,decimals = 2)
```
 
-   $\bar{x}$ = 13.01
 
-   *s* = 0.91
 
-   *n* = 193

An perhaps we want to test the null hypothesis that $\mu_0$ = 13.0. Maybe we have reviewed some scientific literature explaining that this is a standard size for whales, and we are interested to know if our sample matches the value.

The figure below shows the distribution of the sample lengths, with the sample mean in red, and the null hypothesis value in blue.

  
```{r, out.width="80%", echo = FALSE}

Whale_49 %>%
  ggplot(aes(x=length.m.))+
  geom_bar()+
  xlab("Length (m)")+
  ylab("Count")+
  geom_vline(xintercept = 13.01,col="red",linewidth=2)+
    geom_vline(xintercept = 13.0,col="blue",linewidth=2,linetype=2)+
  plot_theme

```

Using any statistical software we can then formulate this as a one sample t test.

```{r}
t.test(Whale_49$length.m.,mu = 13.0)
```


Then, the t-statistic is *t* = 0.133, and the *p-value* = 0.894, so we can't reject the null hypothesis based on this evidence. 

In the tool below you can see what happens to the test, as we change our hypothetical null hypothesis value. The dotted red lines represent the chosen confidence interval around our sample mean.


```{r testUI, context="render",echo=FALSE}
  numericInput("mu_0",label = "Enter Null Hypothesis Mean:",value = 13,min = 10,max=16,step = 0.05)
  selectInput("confidence",label = "Confidence Level",choices = c("90%"=0.95,"95%"=0.975,"99%"=0.995,"99.9%"=0.9995),
              selected=0.975)
plotlyOutput("plot")
```

```{r testSERVER,context="server"}

output$plot<-renderPlotly({
  
  ci<-mean(Whale_49$length.m.)+(c(1,-1)*qt(as.numeric(input$confidence),df=192)*(sd(Whale_49$length.m.)/sqrt(193)))
  
  (Whale_49%>%
  ggplot(aes(x=length.m.))+
  geom_bar()+
  xlab("Length (m)")+
  ylab("Count")+
  geom_vline(xintercept = 13.01,col="red",linewidth=1)+
      geom_vline(xintercept = ci,col="red",linewidth=0.5,linetype=2)+
    geom_vline(xintercept = as.numeric(input$mu_0),col="blue",linewidth=1,linetype=2)+
  plot_theme+
    annotate(geom="text",y=25,x=14.5,label=paste("t =",round(t.test(Whale_49$length.m.,mu = as.numeric(input$mu_0))$statistic,3),
               "\np =",ifelse(round(t.test(Whale_49$length.m.,mu = as.numeric(input$mu_0))$p.value,3)<0.001,
                           "<0.001",
                           format(round(t.test(Whale_49$length.m.,mu = as.numeric(input$mu_0))$p.value,3),nsmall=3))
    )
             )
  )  %>%
    ggplotly()
  
}
  
)

```


If we had tested the following null hypothesis, we would have reached the following results:
 
-   $\mu_{01}$ = 12.90, *p-value* = 0.100, we cannot reject the null hypothesis.
 
-   $\mu_{02}$ = 12.85, *p-value* = 0.017, we we can reject the null hypothesis if we set the significance level at α = 0.05, but we cannot if α = 0.01.
 
-   $\mu_{03}$ = 12.80, *p-value* =0.002, we we can reject the null hypothesis if we set the significance level at α = 0.05 or 0.01, but we cannot if α = 0.001.
 
-   $\mu_{03}$ = 12.75, *p-value* <0.001, we we can reject the null hypothesis for any standard threshold value.

Note that, although the null hypothesis may change, the limits of the confidence interval don't, unless you change the level of confidence.

The t.test is based on some assumptions: 
 
-   the sample mean, $\bar{x}$ is normally distributed. For sample sizes *n > 5* this is usually the case, due to the *Central Limit Theorem*, that states that the sampling distribution of the mean will always be normally distributed, as long as the sample size is large enough, regardless the distribution of the data. As we will have just one sample mean, we will have just to assume that this is the case.
 
-   The individual observations are independent. This is a very strong assumption, and we will find it in most of the other 'simple' hypothesis tests. It basically means that one measurement does not depend on any of the other measurements. When measuring the length (or observing the sex) of a whale, the length or sex of the previously measured/observed whales do not provide any information on this one: if the previous whale was a male, or a long whale does this provide any further clue on what sex or what length will be the current whales? But think about the process of sampling whales across the whole year - there is a good chance we may end up observing the same whale on more than one occasion. If we were able to tag and track the whales we would be confident that this was not happening; but whether this was able to be done in 1949 is perhaps more of a challenge!

Other examples might relate to the concentration of a pollutant downstream is not independent from the concentration of that pollutant upstream, or the temperature today is not independent from yesterday's temperature, or samples being obtained from a clustered process where samples from one cluster will not independent of each other.     

### Non parametric one sample tests

We can test whether the center of a population is a given value, without any assumption on the distribution of the variable or its mean. For each *parametric test* such as the t-test (it is *parametric* because we make assumptions on the population distribution), there is a *non-parametric* test that we can use when we can't (or we don't want to) make any assumption on the population distribution. They can be used always, even when we could use the parametric test such as the t-test, however, their results are usually less conclusive. 

The non-parametric test equivalent to the one-sample t-test is the *Wilcoxon one-sample Signed-Rank test*. It is based on ranks and because of that, the location parameter is not here the mean but the median. This test allows to test the null hypothesis that the *sample median* is equal to a given value:
$$H_0: median = m_0$$
Data is then sorted, and the observations are given a value, either *-1* if if $x < m_0$ or *+1* if $x > m_0$. If the true median is $m_0$ then the sum of this new values should be equal to 0 or very close to it. Based on this, we obtain a p-value and we can make de decision.

We can use the sample we used for the t-test example to see what happens if we used the Wilcoxon one-sample Signed-Rank test:
 
-   $m_0$ = 13.5, *p-value* = 0.2729, we cannot reject the null hypothesis.
 
-   $m_{01}$ = 13.4, *p-value* = 0.9082, we cannot reject the null hypothesis.
 
-   $m_{02}$ = 13.6, *p-value* = 0.04383, we can reject the null hypothesis if we set the significance level at α = 0.05, but we wouldn't if α = 0.01.



### Comparing the mean of two numeric variables: t-test and Wilcoxon Signed-Rank test

To test whether two population means are the same, we are going to use the same tests we used to test one sample's mean: the t-test and the Wilcoxon Signed-Rank test. Testing that two values (the two population means) are the same, is equivalent to test that their difference equals to zero:
$$H_0: \mu_1 - \mu_2 = 0$$, where $\mu_0$ is a numeric value.

Our software of choice will do the hard work for us and we will obtain a p-value that will allow us to make a decision regarding the two population means.

For the parametric test, we will have to add one assumption to those we needed in the one sample test:
 
-   the difference of sample means, $\bar{x_1}-\bar{x_2}$ is normally distributed. 
 
-   The two populations being compared have the same variance. We can test this. there are different tests, we are going to use the *F-test*, see below, although some times a plot could be enough to understand if this assumption holds.
 
-   The individual observations are independent. 

The non-parametric alternative is the Wilcoxon Rank-Sum Test (also known sometimes as the Mann-Whitney U test), and is based on ranking the observations. We will only need to test the assumption of independence to perform this test.

Example: are the male whales longer than female whales? The figure below shows the histogram of the sample lengths for both, male and female.


```{r, out.width="80%", echo = FALSE}
img <- readPNG("img/2Samplet-test.png")
grid.raster(img)
```

 
-   $H_0: \mu_M - \mu_F = 0$ 
 
-   Sample of male whales: $\bar{x_M}$ = 13.093, $s_M$ = 1.33, n=113
 
-   Sample of female whales: $\bar{x_F}$ = 13.561, $s_F$ = 1.51, n=87
 
-   t-test: p-value = 0.02353, we can reject the null hypothesis if we set the significance level at α = 0.05, but we wouldn't if α = 0.01.
 
If we use the Wilcoxon test:  
 
-   The null hypothesis (H0) is that the two populations are equal. 
 
-   The alternative hypothesis (H1) is that the two populations are not equal.

Some researchers interpret this as comparing the medians between the two populations (in contrast, parametric tests compare the means between two independent groups). In general, this is not valid.

-   Wilcoxon test: p-value = 0.04679. In theory, we could reject the null hypothesis if we set the significance level at α = 0.05, but the p-value is very close to the α = 0.05 threshold, so we should be very cautious when rejecting the null hypothesis here.

Therefore, according to the t-test, we could reject the null hypothesis that male and female whales are equally long at the significance level of α = 0.05, but we wouldn't if α = 0.01. According to the wilcoxon test we should be cautious when rejecting the null hypothesis.


### Comparing two variances: F-test

This is a test that we can perform along with the two-sample t-test, to make sure that the assumption of equal variances holds. 

F-test to compare two variances:
 
-   $H_0: \sigma_1^2 = \sigma_2^2 = 0$ 

In the example we used to perform the two-sample t-test, the figure showing the histograms of the male and female sub-samples lengths doesn't show any difference in the variability of the data for the two groups. Let's, however, perform the test:
 
-   $H_0: \sigma_M^2 = \sigma_F^2 = 0$ 
 
-   Sample of male whales: $s_M$ = 1.33, n=113
 
-   Sample of female whales: $s_F$ = 1.51, n=87
 
-   test *p-value* = 0.2022

we cannot reject the null hypothesis that the two variances are equal, so in the t-test we performed above all three assumptions hold.


### Comparing one proportion to a value or two proportions

The tests to compare proportions are based in the fact that we can use the normal approximation. This is not an assumption, it is a fact, so there are no distribution assumptions to be made.

Testing that one population proportion is equal to a value $p_0$, means setting the null hypothesis: $$H_0: p = p_0$$ 

Testing that two population proportions are equal, means setting the null hypothesis: $$H_0: p_A = p_B$$
Example: is the proportion of males = 50%. We use the same sample of 200 whales used to compare lengths and variances, and then perform the test:
 
-   $H_0: p_M = 0.5$

In the sample, there are 113 male and 87 female whales, so the proportion of males is $p_M$ = 0.565. The test p-value is *p-value* = 0.0771 so we cannot reject the null hypothesis that the proportion of whales is 0.5. 

Let's compare now the proportion of male whales in two periods. Is the proportion of caught males in the period 1925-1955 the same as the proportion of caught males in 1956-1978? We are going to use the same sample of 200 whales:
 
-   $H_0: p_{25-55} = p_{56-78} = 0$ 
 
-   period 1925-1955: $n_{25-55}$ = 59, $p_{25-55}$ = 0.567
 
-   period 1956-1978: $n_{56-78}$ = 54, $p_{56-78}$ = 0.468
 
-   test *p-value* = 0.2105

Therefore, we cannot reject the null hypothesis that the proportion of caught males has been the same in the two periods.


### Comparing the mean of more than two numeric variables: ANOVA and Kruskal-Wallis test

To test if *k > 2* population means are the same, we will use the ANOVA and the Kruskal-Wallis test. ANOVA is the acronym of ANalysis Of VAriance, because it compares the variability among the $k$ sample means, one per each group, with the variability within the samples.

for the ANOVA, the null hypothesis to test is:
$$H_0: \mu_A = \mu_B = \mu_C = ... = \mu_K  $$.

Our software of choice will do the work and we will obtain a p-value that will allow us to make a decision regarding the whether the *k* population means are equal or at least one of them is different. 

For the ANOVA, we will have the same assumptions that we had for the 2-sample t-test:
 
-   the sample means, $\bar{x_i}$ are normally distributed. 
 
-   The *k* populations being compared have the same variance. We can test this, however, we will usually perform the test graphically, and, therefore, we won't get any p-value out of it.
 
-   The individual observations are independent. 


Example: we want to test whether the mean length has changed over time. To do so, we are going to split the catches in the following four periods:  *1925-1929*, *1949-1954*"*, *1955-1959*"*, and *1960-1963*. The figure below shows the histogram of the sample lengths for the four periods.


```{r, out.width="80%", echo = FALSE}
img <- readPNG("img/Anova.png")
grid.raster(img)
```

We are going to perform the following test:

-   $H_0: \mu_A = \mu_B = \mu_C = \mu_D$, where the subscripts *A*, *B*, *C* and *D*  refer to the four periods in which we have split the data,
 
-   ANOVA: p-value = $2.607*10^{-08}$, it is very small, therefore we can reject the null hypothesis  at any of the usual significant levels,  α.

According to the plot, the variability of the four samples seems to be quite similar. Since the ANOVA is just a special case of linear model, we can use the tools and methods used in the analysis of linear regression models to test the assumptions for the ANOVA. In Module 4 you will discover those such techniques.

The non-parametric test alternative is the Kruskal-Wallis test. We will only need to test the assumption of independence to perform this test.
 
-   Kruskal-Wallis test: p-value = $2.054*10^{05}$. Again, the p-value is very small and therefore  we can reject the null hypothesis  at any of the usual significant levels,  α.


Both tests, ANOVA and Kruskal-Wallis test the null hypothesis that all the *k* population means are equal. When we reject the null hypothesis we have evidence that not all the means are equal, but this can mean, basically, anything: *k-1* means are not different and 1 mean is different from all the others, or all *k* means are different from one each other, or anything in between. We have tests to better understand what it means to reject the null hypothesis in an ANOVA or Kruskal-Wallis tests, that involve *multiple comparisons*: comparing each mean to all the others. The number of comparisons to perform grows quite quickly with *n*: if *n=3* we need 3 mean comparisons, if *n=4* we will need 6 comparisons, and if *n=5* 10 comparisons are needed.

The p-value is the probability of being wrong when rejecting the null hypothesis, and we usually set a significance level for the p-value as a threshold to reject the null hypothesis. For example, if we set α = 0.05, the probability of detecting a difference just by chance (incorrectly rejecting the null hypothesis) is 5%. However, this is valid when we perform *one single* test. When we compare groups multiple times, the probability of finding at least one difference just by chance increases depending on the number of times we perform the comparison: for *n=4* comparisons, the probability of finding at least one difference just by chance is 26.5% and this probability increases to 40.1% when *n=10*. The multiple comparison problem also applies to confidence intervals: the probability that at least one 95% confidence interval does not contain the true population value when obtaining 10 simultaneous 95% confidence intervals is 40.1%!

[TO BE FINISHED}]



## p-value and sample size

We have already discussed the relationship between the significance level, the p-value and the confidence interval. In Module 2 we discussed the relationship between the width of the confidence interval and the sample size, and we saw that the width of the confidence interval is a function of $1/\sqrt(n)$, where *n* is the sample size. 

Therefore, if there is a true difference (between 2 or between *k* population means, or between the population mean and a value) 
 
-   we are more likely to detect the difference by rejecting the null hypothesis, the larger the sample size is. 
 
-   At the same time, for a given sample size, the smaller the difference we want to make sure we are able to detect is, the more difficult to reject the null hypothesis will be. 
 
-   In addition, the variability of our data also plays a role in the tests: given a sample size and a difference we want to be sure to detect, the more variable our data is the more difficult it will be to reject the null hypothesis. 
 
-   Finally, if all the above are fixed, the minimum sample size depends on the sampling scheme used. It will be smallest for a Simple Random Sampling design, and will be larger for other designs, and how larger depends on the particular design used.

Therefore, we can determine our sample size as the smallest one that will allow us to reject the null hypothesis given the minimum difference we would like to make sure we are able to detect. To do that we need to define this minimum difference to be detected and we need an estimate of the population measure's variance. 

We can also determine the minimum sample size to detect a difference in proportions. The only difference is that the minimum sample size depends on the proportion $p$, being maximal at 0.5 and decreasing as it approaches the 0 or 1 limit. Therefore, we need an initial estimate of the proportion $p$ that we want to estimate and/or test.

Given all the assumptions we need to make, it can be difficult to determine the minimum sample size, and the suggestion is to use all the budget available for data collection.

**[I PREFER NOT TO GO DEEPER ON THIS, AS I WOULD NEED TO INTRODUCE THE POWER OF A TEST, AND TO USE SOME FORMULAS HERE WITH t OR Z VALUES]**









## External Links and Resources [TO BE UPDATED]

Webster, R. and Lark, M. (2013) Field Sampling for Environmental Science and Management
(This book provides guidance on the design and analysis of sampling method, backed by sound rationale and statistical theory. It concentrates on design-based sampling for estimates of mean values of environmental properties, emphasizing replication and randomization.) 

[Populations and samples](https://www.bmj.com/about-bmj/resources-readers/publications/statistics-square-one/3-populations-and-samples)

[Confidence Intervals](https://www.bmj.com/about-bmj/resources-readers/publications/statistics-square-one/4-statements-probability-and-confiden)

[Sampling and Estimation](https://www.youtube.com/watch?v=kEUN7_b-ja8)