---
title: "Module 3 - Hypothesis Testing"
output: 
  learnr::tutorial:
    css: css/style.css
runtime: shiny_prerendered
description: > 
  Hypothesis Testing Module for the 2025 Envision course
editor_options: 
  markdown: 
    wrap: 72
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(shiny)
library(gt)
library(plotly)
library(png)
library(jpeg)
library(grid)

Whale<-read.csv("Data/Humpback.csv") %>% filter(length.m.>0)
Whale$sex<-factor(Whale$sex, levels = c("Male", "Female", "Hermaphrodite", "Unknown"))


choices<-c("Boxplot"="boxplot",
"Histogram"="histogram",
"Density Plot"="density")


plot_theme<-theme_light()+
  theme(plot.title=element_text(face =2,size=18),
        axis.title=element_text(face =2,size=14),
        plot.subtitle=element_text(face =3,size=14),
        axis.text=element_text(face =1,size=12),
        strip.text=element_text(face =1,size=12))


Whale %>%
  ungroup() %>%
      summarise("Number of whales"=n(),
                "Mean length"=mean(length.m.)%>%round(1),
                "Median length"=median(length.m.)%>%round(1),
                "Minimum"=min(length.m.)%>%round(1),
                "Maximum"=max(length.m.)%>%round(1),
                "Range"=(max(length.m.)-min(length.m.))%>%round(1),
                "Standard Deviation"=sd(length.m.)%>%round(1),
                "Variance"=var(length.m.)%>%round(),
                )%>%
  gather(Statistic,Value) ->table_stat
```

## Introduction

In this third workbook, we are going to be taking a closer look at hypothesis testing.
 
Sampling consists of selecting some part of a population to observe, so one may estimate something about the whole population, or infer and test hypothesis about the population characteristics, or fit models that explain or predict some population outcome. In this workbook we are going to discuss some general ideas about sampling and their relationship to estimation. In Workbook 3 we are going to discuss inference and hypothesis testing and in Module 4 and 5 model fitting.
As an example of sampling for estimation, in order to estimate the amount of lichen available as food for caribou, a biologists collects lichen from selected small plots within the study area. Based on the dry weight of the sample collected, the available amount of food for the whole region can be estimated. Similarly, to estimate the average weight of the individuals of a bird species, some individuals are captured, weighted (and released) and based on these measurements, the average weight of the individuals can be estimated.

Sampling is the process used to collect the data, estimation uses the sampled data to provide an estimate of a population parameter using such data.   

In this session we will cover:

-   A brief description of some of the most used types of sampling units in environmental sciences

-   What is sampling and what are the desirable properties of a good sampling scheme

-   The definition and some examples of the two types of errors that we will find when using sampled data to estimate population characteristics

-   The basic sampling scheme, the Simple Random Sampling, the estimation of population characteristics under this scheme and the computation of Confidence Intervals

-   Finally, we will review some of the most commons sampling strategies used in environmental sciences.

In these workbooks all output produced will come from the use of R, and all of the code used in the workbooks can be found on GitHub here: [CHANGE GITHUB REPOSITORY] <https://github.com/stats4sd/Stats_ReIntro_M1>. But the principles are identical regardless of which data analysis tool is being used.Whichever tool is being used for data analysis - it will require that the user is familiar enough with it to be able to generate any sort of summary or plot or transformation that they might suddenly think could be useful!

## Data being used in this workbook
In this session we are going to use a public dataset produced by the Ocean Biodiversity Information system (OBIS). It contains the Historic whale catch records from Australian whaling stations (1925-1978). The full data set and other details can be found here <https://obis.org/dataset/43e4c2cc-5886-427f-a4a9-d47d2635d246>. It contains 38267 records, but we will use a subset: the data corresponding to the most common caught whale, the Humpback (Megaptera novaeangliae). 

We have also edited it to contain only the columns we are interested in. Below you can explore the the data we are going to use in this workbook.

```{r, echo=FALSE}
DT::datatable(Whale)
```



## What is an 'Hypothesis' 

What is the proportion of males among humpback whales? Without knowing too much about whales, one might answer 50%, so half of the whales are male and the other half are female. Actually, one out of the 23,581 whale was labelled as hermaphrodite, but we can ignore this, and assume that the whales are either male or female.
Let's have a look at the data. The figure below shows the percentage of male whales for the period 1949-1962, the only period with data for every year and with at least 100 catches per year.


```{r fig1, out.width="80%", echo = FALSE}
img <- readPNG("img/MalePerc.png")
grid.raster(img)
```

If there were no difference in the underlying proportions of male and female whales, then each year there would be a 50:50 chance of having more male than female catches, or more female than male catches - just like flipping a coin. However, having more male than female every year for fourteen years is like flipping a fair coin 14 times in a row and getting heads every time. The probability of this happening is 0.5^14, which is a small number: 0.0000610.

If we take into account the other four years with more than 100 catches (1925-28), all of them with proportions of males above 0.5, the probability of this happening - if there were a 50:50 chance to have a male or a female whale - would be 0.00000381. If we observed this in a coin-flipping experiment, we could confidently claim that the coin was not fair.

So, we can ask ourselves whether the true proportion of males among the humpback whales can really be 50%? The process we follow to answer a question like this is called *hypothesis testing*.

We can define a hypothesis as a proposed explanation for a phenomenon. It is not the absolute truth, but rather a provisional, working assumption. Sometimes, it is useful to draw an analogy with the suspect in a criminal case.


## The Null Hypothesis

The *null hypothesis* is the hypothesis - a proposed explanation for a phenomenon - that we work with until we have sufficient evidence against it. In the example of the proportion of male whales, we could state the null hypothesis as:

-   The proportion of male whales is 50%, or

-   The proportion of male whales is equal to the proportion of female whales.

The null hypothesis is what we are willing to assume is the case until proven otherwise. It is extremely conservative, denying any progress or change. However, it does not mean that we believe the null hypothesis is literally true; we can never claim that the null hypothesis has actually been proved. 

In the words of Ronald Fisher, the father of modern statistics:

*"the null hypothesis is never proved or established, but is possibly disproved, in the course of experimentation. Every experiment may be said to exist only in order to give the facts a chance of disproving the null hypothesis."*

*(R.A. Fisher (1935), The Design of Experiments. Oliver and Boyd, p.19).*

There is a strong analogy with criminal trials in the English – and indeed many other – legal systems: a defendant can be found guilty, but nobody is ever found innocent, simply not proven guilty. Similarly, we may find that we reject the null hypothesis, but if we don't have enough evidence to do so, it doesn't mean that we can accept it as truth. It is just a working assumption until something better comes along.  


## The p-value

Let's suppose we obtain a sample of *n=200* units from the population of whales and obtain the following proportions:

-   Male: 60.5%

-   Female: 39.5%

and the difference is 21.00%. 

The crucial question is whether this difference is large enough to provide evidence against the null hypothesis that the difference is equal to zero. 

50% is the probability of obtaining heads when tossing a coin. We can then toss the coin 200 times and observe the proportion of heads in the sample. And we could repeat the process many times, say 10,000 times, and see what distribution of proportions of males we obtain. The result we obtained is the one shown in the figure below

```{r fig2, out.width="80%", echo = FALSE}
img <- readPNG("img/PermTest.png")
grid.raster(img)
```

The figure shows that, out of 10,000 simulations of observing how many successes we got in 200 tries, if the probability of success was 0.5, we obtained 121 (200x0.605=121) or more successes in only 15 cases. 

So, is it possible that, if the true probability of male and female whales is equal (0.5), we observed 121 males in a random sample of 200 whales? The answer is YES, it is possible, although very unlikely, as we obtained 121 or more in just 15 out of 10,000, a proportion of 0.15%

Instead of running a simulation of 10,000 runs, we could have systematically worked through all the possible permutations of the whale sample. Each of these would generate an observed proportion of male whales, and plotting these would produce a smoother distribution than just using 10,000 simulations. Unfortunately, the number of permutations is so high that it cannot be obtained with current means within a person's lifetime.

Fortunately, though, we don’t have to perform these calculations, since the probability distribution for the proportions under the null hypothesis is based on what is known as the binomial distribution, which gives the probability of a number of successes out of n trials for a given probability of success — 0.5 in our case.

The figure below shows the probability density function (A) and the cumulative probability distribution function (B). In both cases, the value 121 is highlighted with a vertical line, to show how likely it is to obtain 121 successes or more in 200 trials where the probability of success is 0.5. This probability is 0.11397%.
     

```{r fig3, out.width="80%", echo = FALSE}
img <- readPNG("img/distr and cumul disrt ft.png")
grid.raster(img)
```

This probability is the so-called *p-value*, one of the most prominent concepts in statistics as practised today. A more formal definition of *p-value* is *a p-value is the probability of obtaining a result at least as extreme as the one we observed assuming the null hypothesis is really true*.


## One-sided or two-sided test

As we saw in the previous section, the p-value is the probability of obtaining a result at least as extreme as ours if the null hypothesis were really true. What do we mean by *extreme*? 

The p-value we obtained, 0.11% is *one-tailed* because it only measures how likely is that we have observed such an extreme value (60.5%) in favour of males if the null hypothesis was really true. This p-value corresponds to what we know as *one-sided test*. 

However, if we had observed the same proportion of females, it would also have made us suspect that the null hypothesis does not hold. Therefore, we should also calculate the probability of observing 60.5% of either male or female. This is what is known as the *two-tailed* p-value that corresponds to the *two-sided test*. The two-tail probability is twice the one-sided probability, 0.23%, or 0.0023.

   

```{r, out.width="80%", echo = FALSE}
img <- readPNG("img/2-sided.png")
grid.raster(img)
```


## Statistical significance

In our example of the proportion of male and female whales, we had the following:

-   *Null hypothesis*: the proportion of male whales is 50%.

-   We obtained a simple random sample of *n=200* whales.

-   We observed the sex and *measured the proportion of male whales in the sample: 60.5%*.

-   We obtained the *p-value = 0.0023*: the probability of obtaining a value as extreme as this if the null hypothesis is true. This is the *two-sided test*.

Even if there is only a very small chance that we would have obtained this percentage of percentage of males and females if the null hypothesis was really true, the *p-value* is small enough for us to say that the results are *statistically significant*. In other words, the difference between the population value (the proportion of male in our example) and the null hypothesis is *significantly different from zero*. 

The difference between the proportion of males observed in the sample and the null hypothesis value cannot be explained by random variation alone. It occurred because the population value *is not equal to the null hypothesis*, and the sample data provides the evidence for this. 

Since we have incomplete information about the population — we only have the sample — we can always be wrong when making statements like like the following: *"the proportion of male whales is significantly different from 0.5"*. However, we control the risk of being wrong when rejecting the null hypothesis through the p-value. 

Using the analogy of criminal trials in the English legal system, the sample data provide the evidence against the innocence of the defendant, and p-value is the probability of being wrong if the defendant is declared guilty.

We need to define a *significance level*, denoted by α, which is the critical threshold value for the p-value. Therefore: 

-   if the p-value ≤ α, the result is statistically significant, and we reject the null hypothesis
 
-   if the p-value > α, the result is not statistically significant, and we cannot reject the null hypothesis

The most commonly used significance level (α) is 0.05 or 5%, but in some cases a level of 0.1 (10%) or 0.01 (1%) may be used. Therefore, our result may be significant at α = 0.05 but not at α = 0.01. Alternatively, it may be significant at both levels or at neither. 

However, we must acknowledge that this threshold is arbitrary. For example, if we use α = 0.05, a p-value of 0.0499 is essentially not different from a p-value of 0.0501, but we would make a completely different decision if we strictly followed the significance level rule. Values well above or well below α lead us to a straightforward decision, whereas values close to α usually lead to an inconclusive decision.



## Hypothesis testing

We will use the hypothesis testing for a large number of tests (see below the section on *Statistical tests*), each of which answers a different question. But all of them follow the same steps: 

1-  Set up a question in terms of a null hypothesis that we want to check. We usually use the notation $H_0$ for the null hypothesis. 

2-  Choose a test statistic that estimates something that, if it turns out to be extreme enough, would lead us to doubt the null hypothesis. You will not actually have to find this statistic, as each (standard) statistical test has already defined this test statistic. And you will not have to compute it either, as your statistical software will do this for you. 

3-  Generate the sampling distribution of this tests statistic. if the null hypothesis was true. Again, your statistical software will do this for you.

4-  Obtain the p-value comparing your tests statistic with the sampling distribution. The p-value is the probability of observing such an extreme statistic if the null hypothesis is true. We have to decide if we are going to use the one-sided or two-sided p-value. In most of the tests, we will use the two-sided p-value. Again, your statistical software will do this for you, and by default, it will obtain the two-sided p-value.

5-  Declare if the result is significant it the p-value is below some critical threshold α.

It is important to emphasise that the exact p-value is conditional not only on the truth of the null hypothesis, but also on all other assumptions underlying the statistical model, such as lack of systematic bias, independent observations, etc. We'll discuss some of this assumptions later.


### Confidence Interval for hypothesis testing

Although the confidence intervals is a tool for estimation and not for hypothesis testing, we can use it also for hypothesis testing. Instead of generating the sampling distribution under the null hypothesis for the tests statistic and comparing the tests statistic with this sampling distribution to obtain the p-value (steps 3 & 4 in the hypothesis testing process above), we generate the confidence interval and observe whether the null hypothesis value lies within its limits or outside them. Then, if the null hypothesis value lies within the confidence interval limits, we can't reject it, the results are not significant, whereas if the value lies outside the confidence interval limits, the result is significant and there is sufficient evidence to reject the null hypothesis.

Using a confidence level of 95% and a significance level of 5% will lead to the same results, because they are complementary. And the same happens for a confidence level of 99% and a significance level of 1%. If a 99% confidence interval does not include the null hypothesis, then we can reject it with a p-value < 0.01.


## Statistical tests

In this section, we are going to discuss some standard tests used to answer questions about our data. We are not going to go into the mathematical details, we will focus on the type of test and things to consider when running the test. There are plenty of resources on this tests that you can easily find. 

Before any formal hypothesis test, always look at summary statistics and plots showing the relationship you are interested in.


### Comparing the mean of one numeric variable to a value: t-test and Wilcoxon Signed-Rank test

We will develop this first test a little deeper, so you can check all the steps in the hypothesis testing. You can skip the most mathematical bits form, as they will be performed by your statistical software of choice, although they may help you to better understand the output.

The *t-test* for one variable tests the null hypothesis that the population mean is equal to a value: $$H_0: \mu = \mu_0$$, where $\mu_0$ is a numeric value. 
the test statistic is $$ t=\frac{\bar{x}- \mu_0}{\frac{s}{\sqrt{n}}}$$, where
 
-   $\bar{x}$ is the sample mean
 
-   $\mu_0$ is the null hypothesis value to be tested
 
-    *s* is the sample standard deviation
 
-    *n* is the sample size.

This *t* statistic is compared to its sampling distribution were the null hypothesis true (a *Student's t* with *n-1* degrees of freedom), and a p-value is obtained. Then a decision is made based on the p-value and the level of significance we wish to set. 

Lets see an example, we have a sample of 100 whales from our population fo 23,581 whales. The summary statistics are:
 
-   $\bar{x}$ = 13.413
 
-   *s* = 1.429
 
-   *n* = 100

The figure below shows the histogram of the sample lengths.

  

```{r, out.width="80%", echo = FALSE}
img <- readPNG("img/1sampleHist.png")
grid.raster(img)
```


We want to test 
 
-   $\mu_0$ = 13.5

Then, the t-statistic is *t* = -60738, and the *p-value* = 0.545, so we can't reject the null hypothesis based on this evidence. 

If we had tested the following null hypothesis, we would have reached the following results:
 
-   $\mu_{01}$ = 13.4, *p-value* = 0.9266, we cannot reject the null hypothesis.
 
-   $\mu_{02}$ = 13.6, *p-value* = 0.1942, we cannot reject the null hypothesis if we set the significance level at α = 0.05, but we wouldn't if α = 0.01.
 
-   $\mu_{03}$ = 13.7, *p-value* =  0.04749, we can reject the null hypothesis if we set the significance level at α = 0.05, but we wouldn't if α = 0.01.

The 95% confidence Interval is [13.1296; 13.6968], and the 99% Confidence interval is [13.0379; 13.7885]. You can check which values of the null hypothesis are included within the limits of the Confidence intervals. Note that, although the null hypothesis may change, the limits of the confidence interval don't, unless you change the level of confidence.

The t.test is based on some assumptions: 
 
-   the sample mean, $\bar{x}$ is normally distributed. For sample sizes *n > 5* this is usually the test, due to the *Central Limit Theorem*. As we will have just one sample mean, we will have just to assume that this is the case.
 
-   The individual observations are independent. This is a very strong assumption, and we will find it in many other tests. It basically means that one measurement does not depend on the previous one. When measuring the length (or observing the sex) of a whale, the length or sex of the previously measured/observed whales do not provide any information on this one: if the previous whale was a male, or a long whale does this provide any further clue on what sex or what length will be the current whale's?. However, the concentration of a pollutant downstream is not independent from the concentration of that pollutant upstream, or the temperature today is not independent from yesterday's.     


We can test whether the center of a population is a given value, without any assumption on the distribution of the variable or its mean. The for each *parametric test* such as the t-test (it is *parametric* because we make assumptions on the population distribution), there is a *non-parametric* test that we can use when we can't (or we don't want to) make any assumption on the population distribution. They can be used always, even when we could use the parametric test such as the t-test, however, their results are usually less conclusive. 

The non-parametric test equivalent to the one-sample t-test is the *Wilcoxon one-sample Signed-Rank test*. It is based on ranks and because of that, the location parameter is not here the mean but the median. This test allows to test the null hypothesis that the *sample median* is equal to a given value:
$$H_0: median = m_0$$
Data is then sorted, and the observations are given a value, either *-1* if if $x < m_0$ or *+1* if $x > m_0$. If the true median is $m_0$ then the sum of this new values should be equal to 0 or very close to it. Based on this, we obtain a p-value and we can make de decision.

We can use the sample we used for the t-test example to see what happens if we used the Wilcoxon one-sample Signed-Rank test:
 
-   $m_0$ = 13.5, *p-value* = 0.2729, we cannot reject the null hypothesis.
 
-   $m_{01}$ = 13.4, *p-value* = 0.9082, we cannot reject the null hypothesis.
 
-   $m_{02}$ = 13.6, *p-value* = 0.04383, we can reject the null hypothesis if we set the significance level at α = 0.05, but we wouldn't if α = 0.01.



### Comparing the mean of two numeric variables: t-test and Wilcoxon Signed-Rank test

To test whether two population means are the same, we are going to use the same tests we used to test one sample's mean: the t-test and the Wilcoxon Signed-Rank test. Testing that two values (the two population means) are the same, is equivalent to test that their difference equals to zero:
$$H_0: \mu_1 - \mu_2 = 0$$, where $\mu_0$ is a numeric value.

Our software of choice will do the hard work for us and we will obtain a p-value that will allow us to make a decision regarding the two population means.

For the parametric test, we will have to add one assumption to those we needed in the one sample test:
 
-   the difference of sample means, $\bar{x_1}-bar{x_2}$ is normally distributed. 
 
-   The two populations being compared have the same variance. We can test this. there are different tests, we are going to use the *F-test*, see below, although some times a plot could be enough to understand if this assumption holds.
 
-   The individual observations are independent. 

The non-parametric alternative is the Wilcoxon Rank-Sum Test (also known sometimes as the Mann-Whitney U test), and is based on ranking the observations. We will only need to test the assumption of independence to perform this test.

Example: are the male whales longer than female whales? The figure below shows the histogram of the sample lengths for both, male and female.



```{r, out.width="80%", echo = FALSE}
img <- readPNG("img/2Samplet-test.png")
grid.raster(img)
```

 
-   $H_0: \mu_M - \mu_F = 0$ 
 
-   Sample of male whales: $\bar{x_M}$ = 13.093, $s_M$ = 1.33, n=113
 
-   Sample of female whales: $\bar{x_F}$ = 13.561, $s_F$ = 1.51, n=87
 
-   t-test: p-value = 0.02353, we can reject the null hypothesis if we set the significance level at α = 0.05, but we wouldn't if α = 0.01.
 
-   Wilcoxon test: p-value = 0.04679. In theory, we could reject the null hypothesis if we set the significance level at α = 0.05, but the p-value is very close to the α = 0.05 threshold, so we should be very cautious when rejecting the null hypothesis here.

Therefore, according to the t-test, we could reject the null hypothesis that male and female whales are equally long at the significance level of α = 0.05, but we wouldn't if α = 0.01. According to the wilcoxon test we should be cautious when rejecting the null hypothesis.


### Comparing two variances: F-test

This isa test that we sill usually perform along with the two-sample t-test, to make sure that the assumption of equal variances holds. 

F-test to compare two variances:
 
-   $H_0: \sigma_1^2 = \sigma_2^2 = 0$ 

In the example we used to perform the two-sample t-test, the figure showing the histograms of the male and female sub-samples lengths doesn't show any difference in the variability of the data for the two groups. Let's, however, perform the test:
 
-   $H_0: \sigma_M^2 = \sigma_F^2 = 0$ 
 
-   Sample of male whales: $s_M$ = 1.33, n=113
 
-   Sample of female whales: $s_F$ = 1.51, n=87
 
-   test *p-value* = 0.2022

we cannot reject the null hypothesis that the two variances are equal, so in the t-test we performed above all three assumptions hold.


### Comparing one proportion to a value or two proportions

The tests to compare proportions are based in the fact that we can use the normal approximation. This is not an assumption, it is a fact, so there are no distribution assumptions to be made.

Testing that one population proportion is equal to a value $p_0$, means setting the null hypothesis: $$H_0: p = p_0$$ 

Testing that two population proportions are equal, means setting the null hypothesis: $$H_0: p_A = p_B$$
Example: is the proportion of males = 50%. We use the same sample of 200 whales used to compare lengths and variances, and then perform the test:
 
-   $H_0: p_M = 0.5$

In the sample, there are 113 male and 87 female whales, so the proportion of males is $p_M$ = 0.565. The test p-value is *p-value* = 0.0771 so we cannot reject the null hypothesis that the proportion of whales is 0.5. 

Let's compare now the proportion of male whales in two periods. Is the proportion of caught males in the period 1925-1955 the same as the proportion of caught males in 1956-1978? We are going to use the same sample of 200 whales:
 
-   $H_0: p_{25-55} = p_{56-78} = 0$ 
 
-   period 1925-1955: $n_{25-55}$ = 59, $p_{25-55}$ = 0.567
 
-   period 1956-1978: $n_{56-78}$ = 54, $p_{56-78}$ = 0.468
 
-   test *p-value* = 0.2105

Therefore, we cannot reject the null hypothesis that the proportion of caught males has been the same in the two periods.


### Comparing the mean of more than two numeric variables: ANOVA and Kruskal-Wallis test

To test if *k > 2* population means are the same, we will use the ANOVA and the Kruskal-Wallis test. ANOVA is the acronym of ANalysis Of VAriance, because it compares the variability among the $k$ sample means, one per each group, with the variability within the samples.

for the ANOVA, the null hypothesis to test is:
$$H_0: \mu_A = \mu_B = \mu_C = ... = \mu_K  $$.

Our software of choice will do the work and we will obtain a p-value that will allow us to make a decision regarding the whether the *k* population means are equal or at least one of them is different. 

For the ANOVA, we will have the same assumptions that we had for the 2-sample t-test:
 
-   the sample means, $\bar{x_i}$ are normally distributed. 
 
-   The *k* populations being compared have the same variance. We can test this, however, we will usually perform the test graphically, and, therefore, we won't get any p-value out of it.
 
-   The individual observations are independent. 


Example: we want to test whether the mean length has changed over time. To do so, we are going to split the catches in the following four periods:  *1925-1929*, *1949-1954*"*, *1955-1959*"*, and *1960-1963*. The figure below shows the histogram of the sample lengths for the four periods.


```{r, out.width="80%", echo = FALSE}
img <- readPNG("img/Anova.png")
grid.raster(img)
```

We are going to perform the following test:

-   $H_0: \mu_A = \mu_B = \mu_C = \mu_D $, where the subscripts *A*, *B*, *C* and *D*  refer to the four periods in which we have split the data,
 
-   ANOVA: p-value = $2.607x10^{-08}$, it is very small, therefore we can reject the null hypothesis  at any of the usual significant levels,  α.

According to the plot, the variability of the four samples seems to be quite similar. Since the ANOVA is just a special case of linear model, we can use the tools and methods used in the analysis of linear regression models to test the assumptions for the ANOVA. In Module 4 you will discover those such techniques.

*PENDING: MULTIPLE COMPARISONS*

The non-parametric test alternative is the Kruskal-Wallis test. We will only need to test the assumption of independence to perform this test.

 
-   Kruskal-Wallis test: p-value = $2.054x10^05$. Again, the p-value is very small and therefore  we can reject the null hypothesis  at any of the usual significant levels,  α.


## p-value and sample size

We have already discussed the relationship between the significance level, the p-value and the confidence interval. In Module 2 we discussed the relationship between the width of the confidence interval and the sample size. The width of the confidence interval was a function of $1/\sqrt(n)$, where *n* is the sample size. 

Therefore, if there is a true difference (between 2 or *k* population means, or between the population mean and a value) we are more likely to detect it, by rejecting the null hypothesis, the larger the sample size is. At the same time, for a given sample size, the smaller the difference we want to make sure we are able to detect is, the more difficult to reject the null hypothesis will be. Last, the variability of our data also plays a role in the tests: given a sample size and a difference we want to be sure to detect, the more variable our data is the more difficult it will be to reject the null hypothesis.  

[TO BE FINISHED, but you can give it a go, if you want to!!]

Therefore, we can define our sample size as the one that will allow us to reject the null hypothesis if the   







## External Links and Resources [TO BE UPDATED]

Webster, R. and Lark, M. (2013) Field Sampling for Environmental Science and Management
(This book provides guidance on the design and analysis of sampling method, backed by sound rationale and statistical theory. It concentrates on design-based sampling for estimates of mean values of environmental properties, emphasizing replication and randomization.) 

[Populations and samples](https://www.bmj.com/about-bmj/resources-readers/publications/statistics-square-one/3-populations-and-samples)

[Confidence Intervals](https://www.bmj.com/about-bmj/resources-readers/publications/statistics-square-one/4-statements-probability-and-confiden)

[Sampling and Estimation](https://www.youtube.com/watch?v=kEUN7_b-ja8)